---
title: "Brain Image Analysis"
author: "Maximilian Knoll"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load imaging data 

DICOM files can be imported e.g. in MITK (http://www.mitk.org). MITK allows to 
segment whole brain and tumor volumes and then export of the resulting volumes 
as csv files.

These files have the following composition:
x | y | z | VALUE 
whereas (x,y,z) are the non-standardized coordinates, and VALUE the respective
value for T1ce, T2, .. sequences or CT data.

These files are imported in a first step. Several different variants are
possible:
- Import the raw csv files with usually 4 columns
- Store (x,y,z) coordinates in one single value
- +/- compression (e.g. using gzip)

The encoded coordinates are useful to easily perform set operations, and to
save main memory when performing R analyses. 

An encoded value has the following internal form, where {a,b,c} are the reserved 
bit positions for the {x,y,z} values.

x[x..x]ccccccccccbbbbbbbbbbaaaaaaaaaa

The number of repetitions is determined with the BITSIZE option. For the example 
data, 10 is used. 

```{r}
options("BITSIZE"=10)
```

Two brain datasets are included in this package with their respective 
whole brain volumes and GTVs. 

The exemplary data is stored with an encoded coordinate (COORD). 

Both GTV and whole brain data is loaded.

```{r}
## load GTV data, having only a COORD column (contains x,y,z) vals
gtvFile <- system.file("extdata", "TCGA-76-4931/PXL_GTV_COORD.csv.gz", package = "imageanalysisBrain")
gtv <- read.table(gzfile(gtvFile), sep=",", header=TRUE)
gtvImage <- new("brainImage", gtv, gtvFile, 10) #10: Used Bitsize for encoding

##Load Brain data
brainFile <- system.file("extdata", "TCGA-76-4931/PXL_BRAIN_COORD.csv.gz", package = "imageanalysisBrain")
brain <- read.table(gzfile(brainFile), sep=",", header=TRUE)
brainImage <- new("brainImage", brain, brainFile, 10) #10: Used Bitsize for encoding
```


As different techniques are used for data acquisition, the maximum value in
x dimension (coronar) should be normalized. This is done by calculating a
scaling factor.

```{r}
## calculate x scaling factor 
brainImage <- calcXScalingFactor(brainImage)
```


After importing the data, a center of origin is determined to allow for 
evaluation of distances, e.g. between brain and tumor center (getZeroCoord()).


A dataset containing matching GTV, whole brain, point of origin coordinates 
and scaling factors is then stored as a list entry (allData).

## Image features 

To explore characteristics of the available data, a wide range of image 
features can be extracted from the images.

These can be separated into different categories:

- Distribution characteristics: descriptive values, e.g. mean, sd, ...
- Distances between the Tumor-Center (gtvZero) and the brain center
  of origin (brainZero)
- ...

The feature values are calculated for each allData list element (=dataset
containing GTV + brain data) and is stored in an additional list (imgFeatures). 

Technical differences in acquisitions between different datasets has to be 
compensated for. 

The used heuristics are: 
- Tumor volumes are calculated as proportions of the whole brain volume
- Distances are adjusted with the previously mentioned xfactor which
  normalized the maximal biparietal distance to 1.
  
The psych and e1071 packages provide useful functions for the first point.
  
```{r}
## Store all imaging features
imgFeatures <- list()

# GTV in relation to whole Brain Volume
imgFeatures$gtvVolNorm <- length(gtvImage@coords)/length(brainImage@coords)

# Only GTV dependent measurements
imgFeatures$gtvSD <- sd(gtvImage@measurements)

#Distances between points of origins: GTv / Brain
imgFeatures$koordDistCanberraNorm <- dist(
    rbind(gtvImage@origin*brainImage@xfactor, 
          brainImage@origin*brainImage@xfactor),
    method="canberra")
```


## Extract tissue classes from MRI data

To analyze tumor composition, brain / tumor tissue can be separated 
by intensity into different classes. However, absolute cutoffs,
as can be used for CT images, are not applicable due to a often 
high hetereogeneity scanners / sequences. 

Therefore, different "tissue classes"" are extract by analyzing whole 
brain value to identify cutoffs for these classes.
This allows to quantify the relative contibution of e.g. a hypointense 
class to a given tumor volume / GTV. 

This separation is done by repeatedly drawing of small numbers of voxels
from the whole brain volume. Their resultign distribution is then 
analyzed, and aggregated minimum values are selected as cutoffs. 

The next step is a filtering step which selects only local maxima which 
are neighboured by two larger local maxima.

```{r}
## sample data 
## takes approximately 40 sek for 100.000 iterations with 7 running threads
## on a i7-4820K CPU @ 3.70GHz GenuineIntel GNU/Linux
brainImage@selectedThresholds <- sampleData(brainImage@measurements, iterat=10000)
```


The complete dataset consisting of whole brain and tumor data can 
now be splitted according to the identified cutoffs (subclasses list).
These can then be easily plotted (2D, 3D).

```{r}
##TODO: check if no subclasses could be identified....
subclassesGTV <- getSubclasses(gtvImage, brainImage@selectedThresholds)

subclassesBrain <- getSubclasses(brainImage, brainImage@selectedThresholds)
```


## Plot brain / tumor slices

Single z-slices can now be easily plotted.

```{r}
image(brainImage[,,10])
image(gtvImage[,,10])
```


If dimensions should be preserved to ensure direct comparability 
or overlap of images, an helper function can be used.

```{r}
##Plot brain Slice
dim <- plotZSlice(brainImage, 10, dim=NULL) 

## ... and add GTV slice 
plotZSlice(gtvImage, 10, dim=dim, add=T, col=gplots::bluered(1024)) 
```


Showing only the identified subclass is just as easy (Brain):

```{r}
##Plot brain
dim <- plotZSlice(brainImage, 10, dim=NULL) 

##Plot the first subclass (lowest intensity) proportion of the whole brain
plotZSlice(subclassesBrain[[1]], z, dim, add=T, col=gplots::bluered(1024))
```

and GTV:

```{r}
##Plot brain
imageanalysisBrain::plotZSlice(data, z, minx=minx, maxx=maxx, miny=miny, 
maxy=maxy)

##Plot the second subclass as proportion of the GTV
imageanalysisBrain::plotZSlice(subclassesGTV[[2]], z, minx=minx, maxx=maxx, miny=miny, 
maxy=maxy, add=T, col=gplots::redgreen(1024))
```



