---
title: "Brain Image Analysis"
author: "Maximilian Knoll"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load imaging data 

DICOM files can be imported e.g. in MITK (http://www.mitk.org). MITK allows to 
segment whole brain and tumor volumes and then export of the resulting volumes 
as csv files.

These files have the following composition:
x | y | z | VALUE 
whereas (x,y,z) are the non-standardized coordinates, and VALUE the respective
value for T1ce, T2, .. sequences or CT data.

These files are imported in a first step. Several different variants are
possible:
- Import the raw csv files with usually 4 columns
- Store (x,y,z) coordinates in one single value
- +/- compression (e.g. using gzip)

The encoded coordinates are useful to easily perform set operations, and to
save main memory when performing R analyses.

After importing the data, a center of origin is determined to allow for 
evaluation of distances, e.g. between brain and tumor center (getZeroCoord()).

As different techniques are used for data acquisition, the maximum value in
x dimension (coronar) should be normalized. This is done by calculating a
scaling factor.


```{r}
options("BITSIZE"=10)

## Sample TCIA processed datasets
samples <- c("TCGA-76-4931", "TCGA-76-6285")

## collect all data
allData <- list()
for (i in 1:length(samples)) {
    print(paste("Loading", samples[i], ""))
    ## read GTV data
    if (FALSE) {
        ##variant 1: separate columsn for x,y,z value
        gtvFile <- system.file("extdata", 
                                paste(
                                    samples[i],"/","PXL_GTV.csv", sep=""), 
                                package = "imageanalysisBrain")
        gtv <- data.frame(data.table::fread(gtvFile, sep=",", header=TRUE))
        gtv$COORD <- enc(gtv$x, gtv$y, gtv$z)
        } else {
            ##variant 2: all info encoded in COORD column
            gtvFile <- system.file("extdata", 
                                    paste(
                                        samples[i],"/","PXL_GTV_COORD.csv.gz", 
                                        sep=""),
                                    package = "imageanalysisBrain")
            gtv <- read.table(gzfile(gtvFile), sep=",", header=TRUE)
            gtv$x <- imageanalysisBrain::decX(gtv$COORD)
            gtv$y <- imageanalysisBrain::decY(gtv$COORD)
            gtv$z <- imageanalysisBrain::decZ(gtv$COORD)
            gtv <- gtv[,c(3:5,1:2)]
            }
    
    ## read Brain data
    if (FALSE) {
        brainFile <- system.file("extdata", 
                                    paste(
                                    samples[i],"/","PXL_BRAIN.csv", 
                                    sep=""), 
                                    package = "imageanalysisBrain")
        brain <- data.frame(data.table::fread(brainFile, sep=",", 
    header=TRUE))
        brain$COORD <- enc(brain$x, brain$y, brain$z)
        } else {
            ##variant 2: all info encoded in COORD column
            brainFile <- system.file("extdata", 
                                        paste(
                                            samples[i],
                                            "/",
                                            "PXL_BRAIN_COORD.csv.gz", 
                                            sep=""), 
                                        package = "imageanalysisBrain")
            brain <- read.table(gzfile(brainFile), sep=",", 
        header=TRUE)
            brain$x <- imageanalysisBrain::decX(brain$COORD)
            brain$y <- imageanalysisBrain::decY(brain$COORD)
            brain$z <- imageanalysisBrain::decZ(brain$COORD)
            brain <- brain[,c(3:5,1:2)]
            }
    
    ## calculate x scaling factor 
    xvals <- imageanalysisBrain::getXRange(brain)
    xfactor <- 1/(xvals$xMax - xvals$xMin)
    
    ##Bind together and get Points of  Origins for Brain and Tumor Volumes
    allData[[i]] <- list(name=samples[i],
                        gtv=gtv,
                        brain=brain,
                        gtvZero=imageanalysisBrain::getZeroKoord(gtv,
            draw = FALSE),
                        brainZero=imageanalysisBrain::getZeroKoord(brain, 
            draw = FALSE),
                        factor=xfactor
                        )
    }
```


## Image features 

Several image features and distance metrics between brain and tumor centers 
can be easily calculated. 

```{r}
library(e1071) 
require(pheatmap)

#column of interest (e.g. T1KM, T2, ADC...)
index <- 4

imgFeatures <- list()

##basic features
imgFeatures$gtvVolNorm <- 
    do.call(rbind, 
            lapply(allData, function(x) length(x$gtv$COORD)/
                        length(x$brain$COORD)))
imgFeatures$gtvSD <- 
    do.call(rbind, 
            lapply(allData, function(x) sd(x$gtv[,index])))
imgFeatures$gtvRange <- 
    do.call(rbind, 
            lapply(allData, function(x) max(x$gtv[,index])-min(x$gtv[,index])))
imgFeatures$koordDistCanberraNorm <- 
    do.call(rbind, 
            lapply(allData, function(x) 
                dist(rbind(x$gtvZero[,c("x","y","z")]*x$factor, 
                            x$brainZero[,c("x","y","z")]*x$factor), 
                    method="canberra")))

##additional Features
imgFeatures$gtvMean <- 
    do.call(rbind, lapply(allData, function(x) mean(x$gtv[,index])))
imgFeatures$gtvMedian <- 
    do.call(rbind, lapply(allData, function(x) median(x$gtv[,index])))
imgFeatures$gtvSkewness <- 
    do.call(rbind, lapply(allData, function(x) skewness(x$gtv[,index])))
imgFeatures$gtvKurtosis <- 
    do.call(rbind, lapply(allData, function(x) kurtosis(x$gtv[,index])))
imgFeatures$gtvMax <- 
    do.call(rbind, lapply(allData, function(x) max(x$gtv[,index])))

##additional distances 
imgFeatures$koordDistEuclNorm <- 
    do.call(rbind, 
            lapply(allData, function(x) dist(
                rbind(x$gtvZero[,c("x","y","z")]*x$factor, 
                    x$brainZero[,c("x","y","z")]*x$factor))))
imgFeatures$koordDistManhattanNorm <- 
    do.call(rbind, 
            lapply(allData, function(x) dist(
                rbind(x$gtvZero[,c("x","y","z")]*x$factor, 
                    x$brainZero[,c("x","y","z")]*x$factor), 
                method="manhattan")))

## output
mat <- do.call(cbind, imgFeatures)
rownames(mat) <- unlist(do.call(rbind, allData)[,"name"])
colnames(mat) <- c("gtvVolNorm", "gtvSD", "gtvRange", 
                    "koordDistCanberraNorm", "gtvMean", 
                    "gtvMedian","gtvSkewness", "gtvKurtosis", 
                    "gtvMax", "koordDistEuclNorm", 
                    "koordDistManhattanNorm")
```


## Process data

To analyze tumor composition, image volumes are sampled repeatedly 
(sampleData()). The retrieved data can be used to identify meaningful 
cutoffs for further analyses (determineTissueClasses()).

```{r}
## sample data 
allBrain <- list()
for (i in 1:length(allData)) {
    print(paste("Processing", allData[[i]]$name))
    ## use larger values thaCn 5000
    allBrain[[i]] <- imageanalysisBrain::sampleData(allData[[i]]$brain[,4], 
    iterat=5000)
    }

##extract tissue class cutoffsC
print("Collect cutoffs ...")
collectedCutoffs <- imageanalysisBrain::determineTissueClasses(allBrain)
```


## Plot volumes and slices

Now, whole volumes or subclasses can be visualized as volume (3D), 
or plotted per slice (plotZSlice).  


```{r}
#require(rgl)

brain <- allData[[1]]$brain
gtv <- allData[[1]]$gtv

## Plot brain 
brainSparse <- brain[seq(from=1, to=length(brain[,1]), by=150),]
#plot3d(brainSparse$x, brainSparse$y, brainSparse$z, col="gray", 
#axes=F, ylab="", xlab="", zlab="")

##Plot brain and tumor center of origin
coord <- imageanalysisBrain::getZeroKoord(brain, draw = FALSE)
#plot3d(coord$x, coord$y, coord$z, col="red", size=20, add=T)
coordGTV <- imageanalysisBrain::getZeroKoord(gtv, draw = FALSE, col="red",
sparse = 50)
#plot3d(coordGTV$x, coordGTV$y, coordGTV$z, col="red", size=20, add=T)

## Plot first tissue class of GTV (3d)
subclasses <- imageanalysisBrain::getSubclasses(gtv, collectedCutoffs[[2]])
#plot3d(subclasses[[1]]$x, subclasses[[1]]$y, subclasses[[1]]$z,
#col="royalblue", add=T)

## Plot slices 
data <- brain
minx <- min(data$x, na.rm=TRUE)
maxx <- max(data$x, na.rm=TRUE)
miny <- min(data$y, na.rm=TRUE)
maxy <- max(data$y, na.rm=TRUE)
z <- median(data$z)
## Plot brain
imageanalysisBrain::plotZSlice(data, z, minx=minx, maxx=maxx, miny=miny, 
maxy=maxy)
## Plot again and add subclass
imageanalysisBrain::plotZSlice(data, z, minx=minx, maxx=maxx, miny=miny, 
maxy=maxy)
## Add Subclass within GTV
imageanalysisBrain::plotZSlice(subclasses[[2]], z, col=gplots::greenred(100), 
add=TRUE, minx=minx,maxx=maxx, miny=miny, maxy=maxy)
```



